Making Sequential Programs Deterministic

### 2019-03-26 design.tex
formatting and language

\begin{verbatim}
@PolyDet int foo(@PolyDet int x) { }
void bar(@NonDet int random, @Det int value) {
@NonDet int x = foo(random);      // No error.
@Det int y = foo(value);          // No error.
System.out.println(foo(random));  // Error.
}
\end{verbatim}
In the example above, the method \codeid{foo} is annotated with the \<PolyDet> type qualifier which allows
the caller method \codeid{bar} to invoke it with both \<Det> and \<NonDet> arguments. The checker flags an error 
only when the returned value is used in a way that breaks any deterministic specification (Printing a \<NonDet> value
breaks our determinism guarantees).

@@

With the above rules, the checker framework will automatically annotate a local collection with type \<@NonDet> and its
type arguments will get the type \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
and annotate their type parameters also with \<@NonDet>. So a local list of Strings gets the default annotation of \codeid{@NonDet List<@NonDet String>}. While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
type, it would could in a high number of false positives. In the following code snippet,
\begin{verbatim}
void test(@Det List<@Det String> argList) {
List<String> localList = argList;
}    
\end{verbatim}
the checker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
\codeid{@NonDet List<@NonDet> String} and since collections types are invariant.
We therefore recommend that programmers explicitly annotate
all local variables that are collections to reduce false positives. The same reasoning and recommendation applies to arrays.

The determinism checker has additional type refinement rule for the sorting methods as shown in Figure~\ref{fig-sorting}.

\begin{figure}
    $\infer[\sc{\mathit{LIST \  SORT}}]{\Gamma \vdash x : \mathit{Det \ List}\langle\mathit{Det} \ \beta\rangle}{\Gamma \vdash x : \mathit{OrderNonDet \ List}\langle\mathit{Det} \ \beta\rangle & \mathit{x.sort()}}$

@@

    \caption{Sorting type refinement rules}
    \label{fig-sorting}
\end{figure}

The checker type refines a receiver of type \codeid{@OrderNonDet List} to a \codeid{@Det List} if all the
type argument of this \codeid{List} have the type annotation \codeid{@Det} (Rule LIST SORT in Figure~\ref{fig-sorting}). Similar rules apply for \codeid{Arrays.sort()} and \codeid{Collections.sort()}.
As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{@OrderNonDet List<@Det Set<@Det String>>}, it will
be type refined to \codeid{@Det List<@Det Set<@Det String>>}. However a receiver of type \codeid{@OrderNonDet List<@OrderNonDet Set<@Det String>>} will not be type refined.

>>>>

\begin{verbatim}
@PolyDet int foo(@PolyDet int x) { ... }

void bar(@NonDet int random, @Det int value) {
    @NonDet int x = foo(random);      // No error.
    @Det int y = foo(value);          // No error.
    System.out.println(foo(random));  // Error.
}
\end{verbatim}
In the example above, the method \codeid{foo} is annotated with the \<PolyDet> type qualifier which allows
the caller method \codeid{bar} to invoke it with both \<Det> and \<NonDet> arguments. The checker flags an error 
only when the returned value is used in a way that breaks any deterministic specification (Printing a \<NonDet> value
breaks our determinism guarantees).

@@

With the above rules, the checker framework will automatically annotate a local collection with type \<@NonDet> and its
type arguments will get the type qualifier \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
and annotate their type parameters also with \<@NonDet>. So a local list of Strings gets the default annotation of \codeid{@NonDet List<@NonDet String>}. While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
type, it could result in a high number of false positives. In the following code snippet,
\begin{verbatim}
void test(@Det List<@Det String> argList) {
    List<String> localList = argList;
}    
\end{verbatim}
the checker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
\codeid{@NonDet List<@NonDet> String} and since collections types are invariant.
We therefore recommend that programmers explicitly annotate
all local variables that are collections to reduce false positives. The same reasoning and recommendation applies to arrays.

The determinism checker has additional type refinement rules for the sorting methods as shown in Figure~\ref{fig-sorting}. 
\begin{figure}
    $\infer[\sc{\mathit{LIST \  SORT}}]{\Gamma \vdash x : \mathit{Det \ List}\langle\mathit{Det} \ \beta\rangle}{\Gamma \vdash x : \mathit{OrderNonDet \ List}\langle\mathit{Det} \ \beta\rangle & \mathit{x.sort()}}$

@@

    \caption{Sorting type refinement rules}
    \label{fig-sorting}
\end{figure}
The checker type refines a receiver of type \codeid{@OrderNonDet List} to a \codeid{@Det List} if all the
type argument of this \codeid{List} have the type annotation \codeid{@Det} (Rule LIST SORT in Figure~\ref{fig-sorting}). Similar rules apply for \codeid{Arrays.sort()} and \codeid{Collections.sort()}.
As per these rules, if \codeid{sort()} is called on a receiver of type \codeid{@OrderNonDet List<@Det Set<@Det String>>}, it will
be type refined to \codeid{@Det List<@Det Set<@Det String>>}. However a receiver of type \codeid{@OrderNonDet List<@OrderNonDet Set<@Det String>>} will not be type refined.
